Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps

### 2022-12-08 manuscript/paper_revision_1.Rmd
revision 1 martin done for the time being

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Poisson draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas). Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage. Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely on a single proxy, and are rarely able to infer population size or density in absolute terms. In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future. The Bayesian approach is popular because it offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts. These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE)

@@

In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021]. We believe that these problems cannot be overcome by methodological refinements in this area alone. Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.

@@

1.  **Reliance on single proxy**: Most investigations use only one source of evidence. Although multi-proxy approaches exist, the individual proxies only serve to support each other or the main estimator, without explicitly combining them.
2.  **Uncertainty in measurements**: All archaeological evidence is inherently uncertain which is carried through to derived measurements. However, in most studies, single curves are presented as estimates, and the potential error associated is almost never specified. 
3.  **Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density. This could be absolute, i.e. a numerical estimate of population, or relative, i.e. a means of scaling changes in the proxy value to changes in population. Lack of suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates. In the best cases, there is a qualitative assessment of the informative value of the proxy, not sufficiently accounting for the complex nature of archaeological data.

@@

-   **Limited**: We have only incomplete data, and it is usually not very informative.
-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, these regions are very unevenly distributed over time and space.
-   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
-   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets. Systematic distortions are the rule rather than the exception.
-   **Heterogeneous**: All potential proxies have different spatio-temporal scales, granularity, information value, scales, and data formats.
-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute. The transfer functions linking the proxy data with the desired quantity (population) are unknown.
-   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results.

@@

Many of the problems with archaeological population proxies are shared with contemporary demography. In response, demographers have increasingly turned to Bayesian methods to estimate and forecast contemporary population dynamics. For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data. Bayesian approaches are well suited for limited, unreliable and noisy data. Various data sources, even contradictory data, can be brought into a common framework and used to support one another. These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions). Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty increases, but this does not prevent general modelling and estimation. Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate transfer functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.

This modeling technique can thus be used to join different lines of evidence horizontally and vertically and combine their results into a overall estimate, including an assessment of their reliability: contradicting data lead to a lower overall reliability, while a mutual support to smaller confidence intervals. If there is no systematic bias that affects all data sources to the same extent, this results in the most reliable estimate possible through the most heterogeneous set of data sources.

Bayesian radiocarbon calibration is a similar, well-established application in archaeology, where radiometric uncertainty is modelled based on prior stratigraphic information. More recently, archaeologists have also used Bayesian modelling techniques for testing hypotheses relating to demographic models based on ^14^C data [e.g. @crema2021a]. This approach differs from the one presented here in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step forward to a model-based, scientific analysis. However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: lacking a combination with other indicators, one is limited to the problems and conditions of sum calibration. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of time series.

@@

Our case study targets the period between 6000–1000 BCE. The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the ^14^C proxy. The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.

A large number of different proxies can be integrated into a model of this type, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. Table \@ref(tab:tableproxies) provides a non-exhaustive list. For our case study, we used a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).

| Proxies                                        |
|------------------------------------------------|
| Expert estimates                           |
| Ethnographic Analogies                         |
| Carrying Capacity                              |
| Economic modelling                             |
| Extrapolation of buried individuals            |
| Burial anthropology                            |
| Settlement data, number of houses              |
| Settlement data, settlement size               |
| **Aoristic analysis**                          |
| **Dendro dates**                               |
| Amount of archaeological objects               |
| **Radiocarbon sum calibration**                |
| Estimates based on specific object types       |
| **Human impact from pollen or colluvial data** |
| aDNA based estimates                           |
| ...                                            |

@@

From the Neolithic onwards, known settlement areas in Switzerland concentrate along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers excellent data for demographic estimation, but poses very specific problems for such an undertaking. We have high-resolution information on the temporal sequence of individual lakeside settlements by means of dendro data. In these cases, ^14^C data are not as abundant simply because they are inferior to dendro dating.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.

## Summed radiocarbon

Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today. The logic for this can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time. The expected value per time unit for the number of simultaneously existing sites/settlements is the sum of the individual probabilities for the sites, i.e. the area under the curve of the probabilities of 14C data. Incidentally, the same logic applies to the aoristic sum (shown below). The main problem with this proxy is therefore not so much its statistical nature, but mainly the systematic biases that arise from the conservation and finding practicalities in the research process. To counter these, and to detect and mitigate the systematic biases, hierarchical modelling such as that carried out in this study is appropriate.

@@

The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al. [-@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the sum calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.

@@

@@

We include relative dating information obtained from the heritage authorities of the Swiss cantons (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series. However, it is not dependent on radiocarbon dating and thus it avoids the methodological issues of sum calibration. Data from `r nrow(aorist)` sites were included in the aoristic sum.

@@

We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004; @lechterbeck2014]. Evidence of deforestation can therefore provide indications of population dynamics. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.

@@

Sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps. Population development is simulated in a 'process model' using a Poisson process.

@@

A special class of Bayesian hierarchical models are so-called 'state space models', specifically designed for time series. They follow two principles. First, a hidden or latent process is assumed, representing the state of the variable of interest $z$ through the entire time series. Every state of variable z as $z_t$ at a specific time is bound by a Markov process to the state of variable $z$ at time before, at $t-1$. Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $z$ at time $t$. This implies that a relationship between the individual manifestation of variable $y$ is generated over time via the hidden variable $z$, which is not directly observable.

@@

$$
z_t = \beta z_{t-1} + \epsilon_t
\tag{1}
$$

$$
y_t = \alpha z_t + \eta_t
\tag{2}
$$

"The autocorrelation in the states is captured by the parameter $\beta$. The observations are a function of the states only and the parameter $\alpha$ allows the observation at time t to be a biased estimate of the state at time t." [@auger-methe_guide_2021, 4-5]. Dabei sind $\epsilon$ die process variation, und $\eta$ der observation error, welche mit einer passenden Verteilung modelliert werden können.

Um die Struktur der Abhängigkeiten zu verdeutlichen, kann man das Modell auch in terms of probability distributions darstellen:

$$
f(z_t | z_{t-1}, \theta_p), \quad t = 1, ... , T
\tag{3}
$$

$$
f(y_t | z_{t}, \theta_o), \quad t = 1, ... , T
\tag{4}
$$

Dabei sind f and g are two normal probability density functions, and $\theta_p$ and $\theta_o$ are vectors of parameters associated with each equation. Eq. 3 describes the autocorrelation in state values as a first-order Markov process, and Eq. 4 describes how observations depend simply on the states. This definition also demonstrates that states are random variables and thus that SSMs are a type of hierarchical mode [@auger-methe_guide_2021, 5].

This structure makes these models particularly suitable for demographic reconstruction using archaeological and other data. Population density itself is not directly measurable: all we have at our disposal are observations derived by unknown transfer functions.

Our overall model is broken down into several hierarchically-connected individual elements. The process model represents the demographic development itself, without already being explicitly parameterised with data. Here we assume that the latent variable 'number of sites' ($z$) is strongly autocorrelated across different time periods ($z_t$). The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. The population at time $t$ results from the population at time $t-1$ times a parameter $\beta$, which represents the population change at this time (see Equation 1).

@@

$$
\lambda_t = z_{t-1} e^{\mathbf{\beta^T \delta \mathbf{y}_t}}
\tag{7}
$$

$\mathbf{\beta}$ is a parameter vector of weightings that sums to $1$, a vector transpose is indicated by $^T$, and $y_t$ is a vector of proxy difference values.

Using the logarithm as a link function ensures that $\lambda$, which must always be positive, can also be described by variables that may also be negative. $\beta$ serves as slope factor, as in a normal linear regression. Here, it functions as a scaling factor for the individual proxies. $z_t$ is to be understood as an intercept, representing a baseline when there were no change due to the variables. This is the desired behaviour: $\lambda$ is equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.

Since $\lambda$ and $z$ are essentially in the same range (e.g. if $lambda=1$, the expected value for $z$ would also be 1), $z_{t-1}$ must also be log-transformed for the congruence of both values. Population size $z_t$ as well as population change $\lambda_t$ are time-dependent. At each individual point in time, these variables will take on different values. But we can assume that the population change will not exceed certain limits ($max\_change\_rate$), though it is not possible to specify this at this point.

$$
\begin{aligned}
max\_growth\_rate &\sim Gamma(shape = 5, scale=0.05) \\
z_t/z_{t-1} &< (max\_growth\_rate + 1) \\
z_{t-1}/z_t &< (max\_growth\_rate + 1)
\end{aligned}
\tag{8}
$$
A gamma distribution centres probability in the range $[0,1[$; adding 1 makes this range $[1-2[$. This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model. The estimation of this parameter for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data.

@@

In this initial implementation, the observational model is essentially a Poisson regression, where the proxies are used to inform the change in the number of settlements between time steps. The individual proxies were z-normalised and absolute differences between time steps were then computed. If the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.

$$
\begin{aligned}
y'_t &= \frac{y_t - \bar{y}} {\sigma^2_y}\ |\ \sigma^2_y := Standard\ Deviation  \\
\delta y'_t &= y'_t - y'_{t-1}
\end{aligned}
\tag{9}
$$

@@

```{r popdensplot, fig.cap="Estimate of population density predicted by the model. The four input proxies are also plotted (scaled) for comparison."}
knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
```

>>>>

  Here, we present a Bayesian hierarchical model suite that uses Negative Binomial linear model and state-space representation to produce absolute estimates of past population size and density. At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Negative Binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas). Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage. Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely on a single proxy, and are rarely able to infer population size or density in absolute terms. In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future. Bayesian approach is the natural and principled one for fusing data while appropriately propagating uncertainty. This makes it possible to qualifying the uncertainty and credibility attached to forecasts. These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE) based on an initial value

@@

In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g. @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022]. This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021]. While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems. We believe that these problems cannot be overcome by methodological refinements in this area alone. One of those problems is, that most analysis of this kind rely on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021]. Other data for demographic processes, such as age composition of grave groups [@bocquet_appel_explaining_2008; @kohler_long_2014] exist, as does the use of environmental data [@lechterbeck2014]. The combination of a large number of indicators can compensate for the weaknesses of the individual proxies used and open up completely new possibilities for insights [@wolpert_past_2021, 10]. A Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates. Problems such as statistical dispersion of sum calibration and effects of the calibration curve will be mitigated respectively detected by this combination with indicators that are not subject to these effects. Moreover, in our approach, the mutual evaluation of the proxies in terms of concordance penalises divergent behaviour of an indicator (due to intrinsic biases) by reducing its relevance for the resulting estimate. While this does not eliminate these problems entirely, it does reduce them in a numerically manageable way for large data sets.

@@

1.  **Reliance on single proxy**: Most investigations use only one source of evidence. Although multi-proxy approaches exist, the individual proxies only serve to support each other or the main estimator, without explicitly combining them.
2.  **Uncertainty in measurements**: All archaeological evidence is inherently uncertain which is carried through to derived measurements. However, in most studies, single curves are presented as estimates, and no uncertainty measurement exists.
3.  **Lack of a linking function**: By 'linking function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density. This could be absolute, i.e. a numerical estimate of population, or relative, i.e. a means of scaling changes in the proxy value to changes in population. Lack of suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates. In the best cases, there is a qualitative assessment of the informative value of the proxy, not sufficiently accounting for the complex nature of archaeological data.

Our current implementation, with an abductive understanding of the data and the need to set a determinant initial value for the reconstruction, represents a first step towards a fully developed end-to-end hierachical Bayesian model capable of solving these problems.

@@

-   **Limited**: We have only incomplete data, and it is usually not very informative.
-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, these regions are very unevenly distributed over time and space.
-   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
-   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets. Systematic distortions are the rule rather than the exception.
-   **Heterogeneous**: All potential proxies have different spatio-temporal scales, granularity, information value, scales, and data formats.
-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute. The functions linking the proxy data with the desired quantity (population) are unknown.
-   **Contradictory**: When considering several proxies, differences in linking functions, data quality and noisiness inevitably lead to different results.

@@

Many of the problems with archaeological population proxies are shared with contemporary demography. In response, demographers have increasingly turned to Bayesian methods to estimate and forecast contemporary population dynamics. For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data. Bayesian approaches are well suited for limited, unreliable and noisy data. Various data sources, even contradictory data, can be brought into a common framework and used to support one another. These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions). Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty increases, but this does not prevent general modelling and estimation. Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate linking functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.

This modeling technique can thus be used to join different lines of evidence horizontally and vertically and combine their results into a overall estimate, including an assessment of their reliability: contradicting data lead to a lower overall reliability, while a mutual support to smaller confidence intervals. If there is no systematic bias that affects all data sources to the same extent, this results in the most reliable estimate possible through the most heterogeneous set of data sources. Even though data fusion can be achieved with different techniques, the Bayesian approach seems to us to be excellently suited for our purposes due to its flexibility and robustness.

Bayesian radiocarbon calibration is a similar, well-established application in archaeology, where radiometric uncertainty is modelled based on prior stratigraphic information. More recently, archaeologists have also used Bayesian modelling techniques for testing hypotheses relating to demographic models based on ^14^C data [e.g. @crema2021a]. This approach differs from the one presented here in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only. This is a clear step forward to a model-based, scientific analysis. However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: without a combination with other indicators, one is largely limited to the problems and conditions of summation calibration. However, recent work [@price_end__end_2021; @carleton_groucutt_2020] shows a possible way out here, and might be a promising alternative to abductive approaches like ours.

@@

Our case study targets the period between 6000–1000 BCE. The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches. The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.

A large number of different proxies can be integrated into a model of this type, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. Table \@ref(tab:tableproxies) provides a non-exhaustive list. Furthermore, it is possible to combine analytical deductive approaches with the inductive proxies, such as ethnographic analogies and economic modelling. For our case study, we used a landscape openness indicator; an aoristic sum of typological dated sites; a radiocarbon based settlement count (sum calibration); and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).

| Possible Proxies                               |
|------------------------------------------------|
| Expert estimates                               |
| Extrapolation of buried individuals            |
| Burial anthropology                            |
| Settlement data, number of houses              |
| Settlement data, settlement size               |
| **Aoristic analysis**                          |
| **Dendro dates**                               |
| Amount of archaeological objects               |
| **Radiocarbon sum calibration**                |
| Estimates based on specific object types       |
| **Human impact from pollen or colluvial data** |
| aDNA based estimates                           |
| ...                                            |

@@

From the Neolithic onwards, known settlement areas in Switzerland concentrate along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers on the one hand excellent data for population estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

## Radiocarbon Based Settlement Count

Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today. In our variant of the summation calibration, we sum up the individual data in each site so that each site receives a weight of 1. These individual probabilities are then summed up over all sites, resulting in a radiocarbon based settlement count (RBSC). In many respects, this follows the same logic as generally applied summation calibration, where binning achieves much the same outcome.

The logic for RBSC can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time. The expected value per time unit for the number of simultaneously existing sites/settlements is the sum of the individual probabilities for the sites, i.e. the area under the curve of the probabilities of 14C data. Incidentally, the same logic applies to the aoristic sum (shown below). The main problem with this proxy is therefore not so much its statistical nature, but mainly the systematic biases that arise from the conservation and finding practicalities in the research process. To counter these, and to detect and mitigate the systematic biases, hierarchical modelling such as that carried out in this study is appropriate.

@@

The dataset for the RBSC primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al. [-@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the RBSC, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.

@@

Dendrodated lakeshore settlements and RBSC are two fairly direct proxies for settlement numbers over time. However, both are subject to certain constraints, especially on the Swiss Plateau, but also beyond. They require the discovery of corresponding settlement sites. This is subject to preservation and discovery filters, which distort the results of the derived estimates. These biases are not connected to past population trends, but result from research activity, preservation and publication status, as well as dating strategies. Therefore, in our model we supplement these two direct proxies with such proxies that are more independent of these biases.

@@

To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, which often have a low dating accuracy. We incorporate this data into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided (scatter, calibration curve effects). Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

@@

The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004]. Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021]. In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used. The technical steps are also documented in the accompanying R compendium. The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)). Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species. Cereal pollen was explicitly retained as an important anthropogenic indicator. As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.

@@

RBSC, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies. The choice of interval reflects a trade-off between the data quality of the proxies and the complexity of the model. The aoristic sum was not smoothed, because it already has a very coarse temporal resolution. In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps. Population development is simulated in a 'process model' using a Negative Binomial process.

@@

A special class of Bayesian hierarchical models are so-called 'state space models', specifically designed for time series. They follow two principles. First, a hidden or latent process is assumed, representing the state of the (vector) variable of interest $z$ through the entire time series. Every state of variable z as $z_t$ at a specific time is bound by a Markov process to the state of variable $z$ at time before, at $t-1$. Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $z$ at time $t$. This implies that a relationship between the individual manifestation of variable $y$ is generated over time via the hidden variable $z$, which is not directly observable.

@@

$$
z_t = \mathbf{\beta} z_{t-1} + \epsilon_t
\tag{1}
$$

$$
x_t = \mathbf{\alpha} z_t + \eta_t
\tag{2}
$$

"The autocorrelation in the states is captured by the parameter $\beta$. The observations are a function of the states only and the parameter $\alpha$ allows the observation at time t to be a biased estimate of the state at time t." [@auger-methe_guide_2021, 4-5]. Here $\epsilon$ is the process variation, and $\eta$ the observation error, which can be modelled with a suitable distribution. In our current implementation we ignore $\alpha$ and thus assume that $\mathbf{x}_t$ is an unbiased observation. We compensate for this (see below) by defining $\beta$ as a simplex with sum 1. This leads to the intuition that each of the proxies is seen as more or less good compared to the others, but in their sum they represent the best possible description of $z_t$ (given the available data).

To clarify the structure of the dependencies, the model can also be represented in terms of probability distributions:

$$
f(z_t | z_{t-1}, \mathbf{\theta}_p), \quad t = 1, ... , T
\tag{3}
$$

$$
f(x_t | z_{t}, \mathbf{\theta}_o), \quad t = 1, ... , T
\tag{4}
$$

Here f and g are two probability density functions, and $\theta_p$ and $\theta_o$ are vectors of parameters associated with each equation. Eq. 3 describes the autocorrelation in state values as a first-order Markov process, and Eq. 4 describes how observations depend simply on the states. This definition also demonstrates that states are random variables and thus that SSMs are a type of hierarchical mode [@auger-methe_guide_2021, 5].

This structure makes these models particularly suitable for population reconstruction using archaeological and other data. Population density itself is not directly measurable: all we have at our disposal are observations derived by unknown linking functions.

Our overall model is broken down into several hierarchically-connected individual elements. The process model represents the population development itself, without already being explicitly parameterised with data. Here we assume that the latent variable 'number of sites' ($z$) is strongly autocorrelated across different time periods ($z_t$). The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on. The population at time $t$ results from the population at time $t-1$ times a parameter $\beta$, which represents the population change at this time (see Equation 1).

@@

$$
\lambda_t = z_{t-1} e^{\mathbf{\beta^T \delta \mathbf{x}_t}}
\tag{7}
$$

$\mathbf{\beta}$ is a parameter vector of weightings that sums to $1$, a vector transpose is indicated by $^T$, and $x_t$ is a vector of proxy difference values.

Using a link function ensures that $\lambda$, which must always be positive, can also be described by variables that may also be negative. $\beta$ serves as slope factor, as in a normal linear regression. Here, it functions as a scaling factor for the individual proxies. $z_t$ is to be understood as an intercept, representing a baseline when there were no change due to the variables. This is the desired behaviour: $\lambda$ is equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.

A Poisson distribution has the constraint that the variance is equal to the mean. This distribution may have been empirically very useful for modelling respective processes. However, we were advised in the peer review of the article that this limitation may not be justified. Therefore, we have rebuilt and run the model using a Negative Binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.

A negative binomial distribution can be seen as a generalisation of a Poisson distribution. In this case, the expected value and variance are parameterised separately, which gives the model an additional degree of freedom. This affects the credibility intervals, which become correspondingly wider. A negative binomial distribution is described by a parameter of the probability of success $p$, and a parameter $r$, which indicates the rate. Therefore, the implementation of the process model in this variant looks like this:

$$
z_t \sim NB(r,p_t) \\
p_t = \frac{r}{r+\lambda_t} \\
r \sim halfCauchy(0,25)
\tag{8}
$$

The negative binomial distribution has a variance $\frac{r (1-p)}{p^2}$, whereby the distribution becomes identical to the Poisson distribution in the limit $r \to \inf$. In practise, values above 50 are considered to produce a distribution that approaches a Poisson Distribution. In the model, $r$ is estimated from the data using a half-Cauchy distribution with a location parameter of 25 as prior. This represents a weakly informative prior according to the standard specifications for a scale hyperparameter [@gelman_prior_2006].

Population size $z_t$ as well as population change $\lambda_t$ are time-dependent. At each individual point in time, these variables will take on different values. But we can assume that the population change will not exceed certain limits ($max\_change\_rate$), though it is not possible to specify this at this point.

$$
\begin{aligned}
max\_growth\_rate &\sim Gamma(shape = 5, scale=0.05) \\
z_t/z_{t-1} &< (max\_growth\_rate + 1) \\
z_{t-1}/z_t &< (max\_growth\_rate + 1)
\end{aligned}
\tag{9}
$$
A gamma distribution is definied [0, inf). This parameterisation used here centres the probability mass in the range 0-1; adding 1 makes this range 1-2. This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model. The estimation of this parameter for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data.

@@

In this initial implementation, the observational model is essentially a Negative Binomial regression, where the proxies are used to inform the change in the number of settlements between time steps. The individual proxies were z-normalised and absolute differences between time steps were then computed. If the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.

$$
\begin{aligned}
x'_t &= \frac{x_t - \bar{x}} {\sigma^2_x}\ |\ \sigma^2_x := Standard\ Deviation  \\
\delta x'_t &= x'_t - x'_{t-1}
\end{aligned}
\tag{10}
$$

@@

## Negative Binomial Model

```{r popdensplot, fig.cap="Estimate of population density predicted by the model. The four input proxies are also plotted (scaled) for comparison."}
knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation_neg_binomial.pdf"))
```
